---
title: "An introduction to patternogram"
author: Jakub Nowosad
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let's start by attaching the necessary packages.

```{r setup}
#| message: false
library(patternogram)
library(sf)
library(terra)
```

Next, we load two spatial datasets.
The first one is a spatial vector point dataset with annual average air temperature measurements in Celsius for Spain in 2019.
The second one is a raster dataset with predictors, such as population density (`popdens`), distance to the coast (`coast`), elevation (`dem`), a satellite-based Normalized Difference Vegetation Index (`ndvi`), and annual average composites of the Land Surface Temperature product for day (`lst_day`) and night (`lst_night`).

```{r}
temp_train = read_sf("/vsicurl/https://github.com/Nowosad/IIIRqueR_workshop_materials/raw/refs/heads/main/data/temp_train.gpkg")
plot(temp_train)

predictors = rast("/vsicurl/https://github.com/Nowosad/IIIRqueR_workshop_materials/raw/refs/heads/main/data/predictors.tif")
plot(predictors, axes = FALSE)
```

## Part 1

```{r}
p_tt = patternogram(temp_train)
p_tt
```

```{r}
plot(p_tt)
```

```{r}
p_tt250 = patternogram(temp_train, cutoff = 250000)
p_tt250
plot(p_tt250)
```

## Part 2

```{r}
nx = minmax(predictors)    
pr = (predictors - nx[1, ]) / (nx[2, ] - nx[1, ])
```

```{r}
p_pr = patternogram(pr)
p_pr
plot(p_pr)
```

```{r}
p_pr_dist = patternogram(pr, dist_fun = "manhattan")
plot(p_pr_dist)
```

```{r}
p_pr_np = patternogram(pr, sample_size = 50)
plot(p_pr_np)
```

```{r}
p_pr_l = patternogram(pr, group = TRUE)
plot(p_pr_l)
```

